---
title: "Bregs and Tregs in lymph nodes"
author: "Daniel Spakowicz and Rebecca Hoyd"
date: "6/24/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(FDRsampsize)
library(dplyr)
library(tidyr)
library(ggplot2)
library(viridis)
```

The following are a series of power calculations to support the study of Breg and Treg cells in cancer, citing a question of interest and then an analysis and associated sample size and power to address it.

# Do lymph nodes involved in tumors have more Breg cells than uninvolved lymph nodes?

Important numbers for this calculation:

* A surgically-resected lymph can yield 1e5 - 1e6 lymphocytes
* Of those an estimated 40% are B cells, with 2-5% of those being Bregs.
* This is an intra-individual question, and therefore is expected to have relatively low coefficient of variation CV = 0.3-0.5

```{r}
# Breg cells collected per sample
ebus.lymph <- 1e6
lymph.b.frac <- .4
breg.frac <- .02
breg.col <- ebus.lymph * lymph.b.frac * breg.frac

# Difference in the fraction of Breg cells, negative binomial distribution
n <- seq(2, 25)
log.fc <- seq(0.01, 1, by = 0.01)

# Calculate power 
power.l <- list()
for (i in 1:length(n)) {
  power.l[[as.character(n[i])]] <- 
    power.hart(n = i, 
               alpha = 0.05, 
               log.fc = log.fc,
               mu = rep(breg.col, length(log.fc)), 
               sig = rep(0.4, length(log.fc)))
}

# Convert to long-format data frame
power.df <- 
  power.l %>%
  bind_rows() %>%
  mutate(log.fc = log.fc) %>%
  gather(n, power, -log.fc) %>%
  mutate(n = as.numeric(n)) %>%
  mutate(fold.change = 2^log.fc)

a1.rep <- 
  power.df %>%
  filter(power >= 0.8) %>%
  filter(log.fc == 0.5) %>%
  head(., 1)

# Plot result
power.df %>%
  ggplot(aes(x = n, y = log.fc, z = power)) +
  geom_raster(aes(fill = power)) +
  labs(x = "Sample Size",
       y = "Log Fold Change",
       fill = "Power") +
  theme_minimal() +
  scale_fill_viridis() +
  geom_hline(aes(yintercept = 0.5), linetype = "dotted", alpha = 0.5) +
  geom_vline(xintercept = a1.rep$n, linetype = "dotted", alpha = 0.5) +
  ggsave("figures/breg-diff_lymphnodes.png", height = 3, width = 4)

```

> The Breg cells will be calculated as a fraction of total lymphocytes. As such, power is estimated as a negative binomial distribution Wald test using the FDRsamplesize package in R. A sample size of `r a1.rep$n` is sufficient to detect a `r round(a1.rep$fold.change, 2)` fold change in NKTcell populations with `r round(a1.rep$power, 2) * 100`% power ($\alpha$ = 0.05, coefficient of variation = 0.4). Code to reproduce all simulations and calculations is available at https://github.com/spakowiczlab/nktcan.


# Do lymph nodes involved in tumors have more Treg cells than uninvolved lymph nodes?

Important numbers for this calculation:

* A surgically-resected lymph can yield 1e5 - 1e6 lymphocytes
* Of those an estimated 50% are T cells, with 5-7% of those being Tregs
* This is an intra-individual question, and therefore is expected to have relatively low coefficient of variation CV = 0.3-0.5

```{r}
# Treg cells collected per sample
ebus.lymph <- 1e6
lymph.t.frac <- .5
treg.frac <- .05
treg.col <- ebus.lymph * lymph.t.frac * treg.frac

# Difference in the fraction of Treg cells, negative binomial distribution
n <- seq(2, 25)
log.fc <- seq(0.01, 1, by = 0.01)

# Calculate power 
power.l <- list()
for (i in 1:length(n)) {
  power.l[[as.character(n[i])]] <- 
    power.hart(n = i, 
               alpha = 0.05, 
               log.fc = log.fc,
               mu = rep(treg.col, length(log.fc)), 
               sig = rep(0.4, length(log.fc)))
}

# Convert to long-format data frame
power.df <- 
  power.l %>%
  bind_rows() %>%
  mutate(log.fc = log.fc) %>%
  gather(n, power, -log.fc) %>%
  mutate(n = as.numeric(n)) %>%
  mutate(fold.change = 2^log.fc)

a1.rep <- 
  power.df %>%
  filter(power >= 0.8) %>%
  filter(log.fc == 0.5) %>%
  head(., 1)

# Plot result
power.df %>%
  ggplot(aes(x = n, y = log.fc, z = power)) +
  geom_raster(aes(fill = power)) +
  labs(x = "Sample Size",
       y = "Log Fold Change",
       fill = "Power") +
  theme_minimal() +
  scale_fill_viridis() +
  geom_hline(aes(yintercept = 0.5), linetype = "dotted", alpha = 0.5) +
  geom_vline(xintercept = a1.rep$n, linetype = "dotted", alpha = 0.5) +
  ggsave("figures/treg-diff_lymphnodes.png", height = 3, width = 4)

```

> The Treg cells will be calculated as a fraction of total lymphocytes. As such, power is estimated as a negative binomial distribution Wald test using the FDRsamplesize package in R. A sample size of `r a1.rep$n` is sufficient to detect a `r round(a1.rep$fold.change, 2)` fold change in NKTcell populations with `r round(a1.rep$power, 2) * 100`% power ($\alpha$ = 0.05, coefficient of variation = 0.4). Code to reproduce all simulations and calculations is available at https://github.com/spakowiczlab/nktcan.

