---
title: "nktcan"
author: "Daniel Spakowicz"
date: "5/19/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(FDRsampsize)
library(dplyr)
library(tidyr)
library(ggplot2)
```

The following are a series of power calculations to support the study of NKTcells in cancer, citing a question of interest and then an analysis and associated sample size and power to address it.

# Do lymph nodes involved in tumors have more NKTcells than uninvolved lymph nodes?

Important numbers for this calculation:
* A lymph sampled by endobronchial ultrasound (EBUS) can sample 1e5 - 1e6 lymphocytes
** of those an estimated 0.2% are NKTcells
* This is an intra-individual question, and therefore is expected to have relatively low coefficient of variation CV = 0.3-0.5

## Do lymph nodes involved in tumors have a different composition of NKTcells than uninvolved lymph nodes?
* NKTcells come in several subtypes, roughly <= 5 (e.g type I and type II) are expected
 
```{r}
# NKTcells collected per sample
ebus.lymph <- 1e6
lymph.nkt.frac <- 0.002
nkt.col <- ebus.lymph * lymph.nkt.frac

n.groups <- 2

# Difference in the fraction of NKTcells, negative binomial distribution
n <- seq(2, 25)
log.fc <- seq(0.01, 1, by = 0.01)

power.l <- list()
for (i in 1:length(n)) {
  power.l[[as.character(n[i])]] <- 
    power.hart(n = i, 
               alpha = 0.05, 
               log.fc = log.fc,
               mu = rep(nkt.col, length(log.fc)), 
               sig = rep(0.5, length(log.fc)))
}

power.df <- 
  power.l %>%
  bind_rows() %>%
  mutate(log.fc = log.fc) %>%
  gather(n, power, -log.fc) %>%
  mutate(n = as.numeric(n)) %>%
  mutate(fold.change = 10^log.fc)

power.df %>%
  ggplot(aes(x = n, y = log.fc, z = power)) +
  geom_contour_filled() +
  labs(x = "Sample Size",
       y = "Log Fold Change",
       fill = "Power") +
  theme_minimal() +
  geom_hline(aes(yintercept = 0.5), linetype = "dotted", alpha = 0.5) +
  ggsave("figures/nkt-diff_lymphnodes.png", height = 3, width = 4)

```

```{r}
i <- 3
    power.hart(n = i, 
               alpha = 0.05, 
               log.fc = log.fc,
               mu = rep(nkt.col, length(fold.change)), 
               sig = rep(0.5, length(fold.change)))
 
    n.hart=2*(qnorm(0.975)+qnorm(0.9))^2*(1/20+0.6^2)/(log(2)^2) # Equation 6 of Hart et al
 power.hart(n.hart,
            0.05,
            log(2),
            20,
            0.6)                        # Recapitulate 90% power  
 res=fdr.sampsize(fdr=0.1,
                  ave.pow=0.8,
                  pow.func=power.hart,
                  eff.size=rep(c(log(2),0),c(100,900)),
                  null.effect=0,
                  mu=5,
                  sig=1)
 res
```

A sample size of `ceiling(n)` is sufficient to detect a change of 1 standard deviation in NKTcell populations with 80% power (\alpha = .

## Do lymph nodes involved in tumors have a different composition of NKTcells than uninvolved lymph nodes?

* NKTcells come in several subtypes, roughly <= 5 (e.g type I and type II) are expected

What fold change can be detected in the subtypes given that less is known about their abundances and distributions? 

```{r}
# Define the false positive rate
sig.level <- 0.05
num.tests <- 5
typeIerror <- sig.level / num.tests

# Sample size
n.total <- 13

# Define ranges of fold changes and coefficients of variation 
# to calculate the power
fold.change <- seq(1, 3, by = 0.01)
coef.variation <- seq(0.1, 1, by = 0.01)

power.l <- list()
for (c in 1:length(coef.variation)) {
  power.l[[as.character(coef.variation[c])]] <- 
    power.hart(n = n.total, 
               alpha = typeIerror, 
               log.fc = log(fold.change),
               mu = rep((nkt.col / num.tests), length(fold.change)), 
               sig = rep(coef.variation[c], length(fold.change)))
}

# Reformat the list for plotting
power.df <- bind_rows(power.l)
power.df$fold.change <- fold.change

# Plot as contour
power.df %>%
  gather(CV, Power, -fold.change) %>%
  mutate(CV = as.numeric(CV)) %>%
  ggplot(aes(x = fold.change, y = CV, z = Power)) +
  geom_contour_filled() +
  labs(x = "Fold Change",
       y = "Coefficient of Variation",
       fill = "Power") +
  theme_minimal() +
  geom_hline(aes(yintercept = 0.5), linetype = "dotted", alpha = 0.5) +
  ggsave("../figures/nkt-subtypes-in-lymphnodes.png", height = 3, width = 4)
```

# Do cancer patients have increased numbers of circulating NKTcells?

There are three cohorts that could be used to address this question at OSU
1) Cancer patients (collected under OSU-?)
2) No cancer, but at high risk (BE WELL Study participants (NCT04267874))
3) Healthy volunteers (collected under OSU-?)
 
* 30-90 million PBMCs from 12 mL blood
* 80-90% of PBMCs are lymphocytes
* 1-2% of lymphocytes are NKTcells

```{r}
pbmcs.per.ml <- 90e6 / 12



```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
